---
title: "NYPD Shooting Incident Analysis (2006-2024)"
author: "Brittany Shaw"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

## Introduction

This report analyzes archived NYPD shooting incident data from 2006 to 2024. The goals are to:

- Understand how shooting frequency varies by **time of day** and **borough**
- Identify **seasonal and monthly trends** in shooting incidents
- Build a **linear regression model** to predict shooting counts based on location, time of day, and month/year

---

## Data Loading and Structure

```{r load-data}
nypd_raw <- read.csv('/Users/brittanyashaw/Downloads/ARCHIVED_NYPD_Shooting_Incident_Data__Historic_.csv')

cat("Rows:", nrow(nypd_raw), "\n")
cat("Columns:", ncol(nypd_raw), "\n")
```

---

## Data Preparation

Shooting times are recorded in mixed formats (e.g., `"0:00:00"` and `"01:00:00"`). The hour is extracted by splitting on `":"` and taking the first element, which handles both formats correctly.

```{r wrangle}
filtered_nypd_df <- nypd_raw %>%
  rename(Borough = BORO, Date = OCCUR_DATE) %>%
  mutate(
    Hour = as.integer(sapply(strsplit(OCCUR_TIME, ":"), function(x) x[1])),
    time_of_day = case_when(
      Hour >= 0 & Hour <= 6  ~ "Overnight",
      Hour >  6 & Hour <= 12 ~ "Morning",
      Hour > 12 & Hour <= 18 ~ "Afternoon",
      Hour > 18 & Hour <= 23 ~ "Evening"
    ),
    Month = as.integer(format(as.Date(Date, format = "%m/%d/%Y"), "%m")),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5)  ~ "Spring",
      Month %in% c(6, 7, 8)  ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Autumn"
    ),
    month_year = format(as.Date(Date, format = "%m/%d/%Y"), "%m-%Y")
  ) %>%
  group_by(Borough, time_of_day, Season, month_year, Month) %>%
  summarise(Shooting_Count = n(), .groups = "drop")
```

---

## Shootings by Time of Day and Borough

The stacked bar chart below shows total shooting counts for each time-of-day window, broken out by borough across the full 2006–2024 period.

```{r plot-time-borough, fig.width=8, fig.height=5}
ggplot(data = filtered_nypd_df,
       aes(x = time_of_day, y = Shooting_Count, fill = Borough)) +
  geom_bar(stat = "identity") +
  labs(
    x     = "Time of Day",
    y     = "Shooting Count",
    title = "Shootings by Time of Day and Borough, 2006–2024"
  ) +
  theme_minimal()
```

**Key observation:** Overnight hours tend to have the highest concentration of incidents, with Brooklyn and the Bronx accounting for a disproportionately large share. Of note, while Overnight hours are the highest, the Evening hours follow very closely behind. 

---

## Average Monthly Shootings

To remove year-to-year noise and surface seasonal patterns, shooting counts are averaged across all years for each calendar month.

```{r avg-monthly}
nypd_avg_monthly <- nypd_raw %>%
  mutate(
    Date  = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
    Month = as.integer(format(Date, "%m")),
    Year  = as.integer(format(Date, "%Y"))
  ) %>%
  group_by(Year, Month) %>%
  summarise(Shooting_Count = n(), .groups = "drop") %>%
  group_by(Month) %>%
  summarise(Avg_Shootings = mean(Shooting_Count))
```

```{r plot-avg-monthly, fig.width=8, fig.height=5}
ggplot(data = nypd_avg_monthly, aes(x = Month, y = Avg_Shootings)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Average Monthly Shootings in NYC (2006–2024)",
    x     = "Month",
    y     = "Average Shooting Count"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal()
```

**Key observation:** Shooting incidents peak in the summer months (June–August) and drop sharply in winter, consistent with seasonal crime patterns reported nationally.

---

## Linear Regression Model

A linear regression model is fit to predict monthly shooting counts using **borough** and **time of day** as predictors. Categorical variables are converted to factors before modeling.

```{r model}
filtered_nypd_df <- filtered_nypd_df %>%
  mutate(
    time_of_day = as.factor(time_of_day),
    Borough     = as.factor(Borough),
    Season      = as.factor(Season)
  )

model <- lm(Shooting_Count ~ Borough + time_of_day,
            data = filtered_nypd_df)

summary(model)
```


**Interpretation notes:**

- **Borough** coefficients indicate the baseline shooting rate difference relative to the reference borough, The Bronx.
- **Time of day** coefficients show the additive effect of each window relative to the reference level, Afternoon.


## Bias and Limitations

As a resident of New York City, it is possible that I may carry notions regarding which boroughs or time periods are more dangerous that could influence how I interpret these findings. Additionally, this dataset only captures reported incidents. Unreported shootings introduce measurement bias, and policing intensity varies by borough, which may affect reporting rates.

---

## Conclusion

This analysis finds that:

1. **Overnight hours and summer months** are associated with the highest shooting volumes.
2. **Brooklyn and the Bronx** consistently account for the largest share of incidents across all time windows.
3. The linear regression model confirms that borough and time of day are both statistically meaningful predictors of shooting count.

Future work could incorporate socioeconomic covariates, spatial clustering, or a time-series model to better capture the autocorrelated structure of monthly counts.

---

*Data source: NYPD Archived Shooting Incident Data (Historic), NYC Open Data.*
